Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-11-03 paper.tex
.

In our approach, we have introduced a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. 
To allow for the exploration of the relative effects of model parameters and spatial configuration, we clustered initial spatial conditions  to represent types of spaces ex-ante (for example: monocentric or polycentric density grids). The sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types. It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.

In order to test for the influence of spatial initial conditions on model outputs, we use a systematic method to compare phase diagrams. We define a phase diagram as the projection of a vector of model outputs in the parameter space. Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values as the median of the final values of an output indicator obtained for the replications of the model initialized with this combination of parameter.\\

Comparing phase diagrams is useful because it allows us to compare the 
Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic (with around 50 different spatial configurations in our applications). A solution is to use systematic quantitative procedures. To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature. We propose to use a relative distance between two phase diagram corresponding to the share of between-diagrams variability relative to their internal variability, given formally by

@@

where $f_{\vec{\alpha}}(\vec{x})$ is the phase diagrams on parameters $\vec{x}$ at given meta-parameters $\vec{\alpha}$. $d$ is a functional distance that we take simply as Euclidian distance. The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.

The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMole~\citep{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file. In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modelling claimed by \citet{cottineau2015modular}.
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructing a spatial generator of density grids}

Our spatial generator applies an urban morphogenesis model \cite{Batty2007}, which has been generalised, explored and calibrated \cite{2017arXiv170806743R}. In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $k$ is reached. The grid then has a population of $k \times N$ (the population lost due to diffusion process to the outside is reallowed through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf. figure \ref{fig:spatialGen}). Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n). Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$? au pas de temps $t$? Pourquoi ne pas indexer alpha? Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.} \comment[FL]{je crois que N c'est l'incrément de pop et n le nombre de fois où c'est smoothé mais c'est quoi la pop totale?}

>>>>

In our approach, we introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. 
To allow for the exploration of the relative effects of model parameters and spatial configuration, we cluster initial spatial conditions  to represent types of spaces ex-ante (for example: monocentric or polycentric density grids). The sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types. It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.

In order to test for the influence of spatial initial conditions on model outputs, we use a systematic method to compare phase diagrams. We define a phase diagram as the vector of model outputs considered as a function of model parameters. Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values as the median of the final values of an output indicator obtained for the replications of the model initialized with this combination of parameters.\\

%Comparing phase diagrams is useful because it allows us to compare the 
Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic (with around 50 different spatial configurations in our applications). A solution is to use systematic quantitative procedures. To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature. We propose to use a relative distance between two phase diagram corresponding to the share of between-diagrams variability relative to their internal variability, given formally by

@@

where $f_{\vec{\alpha_i}}$ are the phase diagrams at given meta-parameters $\vec{\alpha_i}$. $d$ is a functional distance that we take simply as Euclidian distance. The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.

The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMole~\citep{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file. In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modelling claimed by \citet{cottineau2015modular}.
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial generator of density grids}

The spatial generator applies an urban morphogenesis model \citep{Batty2007}, which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}. An open implementation and a characterisation of urban forms the model can produce are provided, what allows us to integrate it easily in our workflow. In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $t_f$ is reached. The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment. We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$. Typical value ranges for the  parameters will be taken as, following \citet{2017arXiv170806743R}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$. We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configuration that can be generated.

%\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n). Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$? au pas de temps $t$? Pourquoi ne pas indexer alpha? Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.} \comment[FL]{je crois que N c'est l'incrément de pop et n le nombre de fois où c'est smoothé mais c'est quoi la pop totale?}
